---
title: "Workshop - Working with DNA sequences and features in R with Bioconductor - version 2"
output:
  html_document:
    number_sections: true
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cerulean
    css: style.css
editor_options: 
  chunk_output_type: console
---

# Installing Bioconductor packages

Bioconductor packages are installed using the CRAN package "BiocManager".

```{r eval=FALSE}
## 1. Install BiocManager to your hard disk:
#
# install.packages("BiocManager")
#
## 2. Use BiocManager to install a BioConductor package to your hard disk:
#
# BiocManager::install("Biostrings")
#
## 3. Load the package into memory in your current R session:
#
# library(Biostrings)
#
```

To update any out-of-date packages use `install()` with no arguments. Restart your R session after doing this!

```{r eval=FALSE}
# BiocManager::install()
```

Bioconductor packages usually have useful documentation in the form of "vignettes". These are readable on the Bioconductor website https://bioconductor.org/packages/release/BiocViews.html#___Software , from the RStudio "help" pane, or within R:

```{r eval=FALSE}
help(package="Biostrings")
# -> select "User guides, package vignettes, and other documentation"

vignette("BiostringsQuickOverview", package="Biostrings")
```


# Main packages used in this workshop

```{r warning=FALSE, message=FALSE}
library(Biostrings)      # Provides DNAString, DNAStringSet, etc
library(BSgenome)        # Provides getSeq()
library(GenomicRanges)   # Provides GRanges containing genomic ranges
library(rtracklayer)     # Provides import() and export()
library(Gviz)            # Provides plotting of genomic features
```


# Sequences

## DNAString

Package `Biostrings` offers classes for storing DNA strings, `DNAString`, amino acid sequences, `AAString`, or anything else in a `BString`. These are like character strings, but a variety of biologically meaningful functions can be applied to them.

```{r}
myseq <- DNAString("CCGCGCACCAAC")
myseq

class(myseq)

reverseComplement(myseq)
translate(myseq)

subseq(myseq, 3,5)
myseq[3:5]

as.character(myseq)
```


You can see a complete set of functions that work with DNAString with:

```{r eval=FALSE}
methods(class="DNAString")
```

You can see internal details of a class with:

```{r eval=FALSE}
getClass("DNAString")
```

The "slots", used to store data internally in these objects, are accessible with, eg, `myseq@length`. However, wherever possible, we should leave the slots alone and use "methods" to access information in the object, eg `length(myseq)`! The classes that this class "extends" tell us some further ways we can interact with it.

You can get help on the `DNAString` class as below. Try this with classes it extends too!

```{r eval=FALSE}
?"DNAString-class"
```


## DNAStringSet

Often we want to work with a list of sequences, such as chromosomes.

```{r}
myset <- DNAStringSet( list(chrI=myseq, chrII=DNAString("ACGTACGT")) )
myset

# A DNAStringSet is list-like
myset$chrII
# or myset[["chrII"]]
# or myset[[2]]
```


## Challenge: sequences {.challenge}

1. Reverse complement the following DNA sequence. Translate the result to an amino acid sequence.

```
"GTAGAGTAATATGGA"
```

2. Also translate bases 3 to 11 of the reverse complement.


# Genomic ranges

We may then wish to refer to regions of these sequences, often with an associated strand. This is done with the `GRanges` type. `GRanges` builds on `IRanges`, "integer ranges". An IRanges has starts and ends. A GRanges additionally has sequence names and strand information.

```{r}
range1 <- GRanges("chrI", IRanges(start=3,end=5), "+")
range1
getSeq(myset, range1)

range2 <- GRanges("chrI", IRanges(start=3,end=5), "-")
getSeq(myset, range2)
```

Accessing GRanges data:

```{r}
seqnames(range1)
start(range1)
end(range1)
width(range1)
strand(range1)
as.data.frame(range1)
```

Internals of an "S4" object such as a GRanges can be accessed using `@`, but this is discouraged. It is better to use the accessor functions above. Observe the completions RStudio offers when you type `range1@`.

```{r eval=FALSE}
# Look at completions for
# range1@
```

Further manipulations:

```{r}
# GRanges are like vectors:
c(range1, range2)

# GRanges can have metadata columns, and are often used like data frames:
mcols(range1)$wobble <- 10
range1
mcols(range1)
range1$wobble

# A handy way to create a GRanges
as("chrI:3-5:+", "GRanges")
```


## Question {.challenge}

Based on what we saw for `DNAString`, how can we learn more about using `GRanges` and `IRanges` objects?


## Visualization

A couple of options for showing ranges are the `GViz` and `ggbio` packages. `ggbio` is a `ggplot2` extension. `ggbio` is a great idea, but (as at late 2021) the package we could use some love to polish and update it. `GViz` takes a `grid` graphics approach. 

Here we'll use the `GViz` package, which is somewhat confusing to use but produces nice graphics out of the box.

```{r fig.height=2}
options(ucscChromosomeNames=FALSE)
look <- function(features, labels=seq_along(features)) {
    axis_track <- GenomeAxisTrack()
    feature_track <- AnnotationTrack(
        features, id=labels, name="", 
        showFeatureId=TRUE, fontcolor.feature="black")
    plotTracks(
        list(axis_track, feature_track), 
        extend.left=1, extend.right=2)
}

look( c(range1, range2) )
```


# Loading files

## Loading sequences

DNA sequences are generally stored in FASTA format, a simple text format. These can be loaded with `readDNAStringSet` from `Biostrings`. Let's load the genome of the *Caeonrhabditis elegans* worm, obtained from the Ensembl FTP site. This is the "soft masked" (dna_sm) version, with repetitive regions given in lower case, but we won't be making use of this today.

```
### After "gunzip"ing, the start of the .fa file looks like this:
# >I dna_sm:chromosome chromosome:WBcel235:I:1:15072434:1 REF
# gcctaagcctaagcctaagcctaagcctaagcctaagcctaagcctaagcctaagcctaa
# gcctaagcctaagcctaagcctaagcctaagcctaagcctaagcctaagcctaagcctaa
# gcctaagcctaagcctaagcctaagcctaagcctaagcctaagcctaagcctaagcctaa
# ...
# ttttttagaaaaattatttttaagaatttttcattttAGGAATATTGTTATTTCAGAAAA
# TAGCTAAATGTGATTTCTGTAATTTTGCCTGCCAAATTCGTGAAATGCAATAAAAATCTA
# ATATCCCTCATCAGTGCGATTTCCGAATCAGTATATTTTTACGTAATAGCTTCTTTGACA
# TCAATAAGTATTTGCCTATATGACTTTAGACTTGAAATTGGCTATTAATGCCAATTTCAT
# ...
```

```{r}
seqs <- readDNAStringSet("Caenorhabditis_elegans.WBcel235.dna_sm.toplevel.fa.gz")
seqs

# Our chromosome name is too verbose.
# Remove everything from the name after the first space.
names(seqs)
names(seqs) <- sub(" .*","",names(seqs))
names(seqs)
```

Conversely, a DNAStringSet can be written to a file with `writeXStringSet`.


## Loading features

Genome annotations are available in a variety of text formats such as GFF3 and GTF. They can be loaded with the `import` function from `rtracklayer`. This GTF file is also from Ensembl, and gives the locations of the genes in the genome, and features within them.

```
### TSV with columns: seqname, source, start, end, score, strand, frame, attributes 
### The start of the .gtf file looks like this:
# #!genome-build WBcel235
# #!genome-version WBcel235
# #!genome-date 2012-12
# #!genome-build-accession GCA_000002985.3
# #!genebuild-last-updated 2019-01
# V       WormBase        gene    9244402 9246360 .       -       .       gene_id "WBGene00000003"; gene_name "aat-2"; gene_source "WormBase"; gene_biotype "protein_coding";
# V       WormBase        transcript      9244402 9246360 .       -       .       gene_id "WBGene00000003"; transcript_id "F07C3.7.1"; gene_name "aat-2"; gene_source "WormBase"; gene_biotype "protein_coding"; transcript_source "WormBase"; transcript_biotype "protein_coding";
# V       WormBase        exon    9246080 9246360 .       -       .       gene_id "WBGene00000003"; transcript_id "F07C3.7.1"; exon_number "1"; gene_name "aat-2"; gene_source "WormBase"; gene_biotype "protein_coding"; transcript_source "WormBase"; transcript_biotype "protein_coding"; exon_id "F07C3.7.1.e1";
# V       WormBase        CDS     9246080 9246352 .       -       0       gene_id "WBGene00000003"; transcript_id "F07C3.7.1"; exon_number "1"; gene_name "aat-2"; gene_source "WormBase"; gene_biotype "protein_coding"; transcript_source "WormBase"; transcript_biotype "protein_coding"; protein_id "F07C3.7.1";
# V       WormBase        start_codon     9246350 9246352 .       -       0       gene_id "WBGene00000003"; transcript_id "F07C3.7.1"; exon_number "1"; gene_name "aat-2"; gene_source "WormBase"; gene_biotype "protein_coding"; transcript_source "WormBase"; transcript_biotype "protein_coding";
# ...
```


```{r warning=FALSE, message=FALSE}
features <- import("Caenorhabditis_elegans.WBcel235.104.gtf.gz")

# Optional: just retain the columns of metadata we need
mcols(features) <- mcols(features)[,c("type","gene_name","gene_id","transcript_biotype","transcript_id")]

features
```


# Seqinfo and genome versions

Various objects have associated Seqinfo information, listing the chromosomes they refer to and their lengths. This allows some extra sanity checking, and is also necessary for some tasks.

```{r}
seqinfo(features)
seqinfo(seqs)

myseqinfo <- seqinfo(seqs)
isCircular(myseqinfo) <- c(rep(FALSE,6),TRUE)

seqinfo(features) <- myseqinfo
```

One thing you need to care about is the genome version. We often refer to genome versions by their name in the [UCSC genome browser](https://genome.ucsc.edu/). For example the two most recent versions of the worm genome are "ce10" from 2010 and "ce11" from 2013. Similarly with human data you may find data for either the "hg19" or "hg38" versions of the human genome.

```{r cache=TRUE}
Seqinfo(genome="ce10")
Seqinfo(genome="ce11")
```

Notice the slightly different chromosome lengths! Features from one genome version will not be correctly located on a different genome version.

Our ENSEMBL sequences match "ce11", also called "WBcel235". However the chromosome names are slightly different, a common source of pain. See `?seqlevelsStyle()`.



# Querying GRanges objects

The metadata columns let us query the GRanges, for example for a feature type:

```{r results="hold"}
cds <- subset(features, type == "CDS")
cds
# Equivalently:
#   features[features$type == "CDS"]
```

Note: `subset` is a generic R function. It is similar to dplyr's `filter`. The second argument is special, in it you can refer to columns of the GRanges directly.


# Representation of genes in a GTF file

Let's have a look at the different feature types in a particular gene.

```{r fig.height=3}
trx1_features <- subset(features, gene_name == "trx-1")
# Equivalently:
#   features[features$gene_name == "trx-1" & !is.na(features$gene_name)]

trx1_features
look(trx1_features, trx1_features$type)
```

Look at the different types in the "type" column. Each "gene" may have multiple "transcript" features (isoforms). Each transcript in turn has a set of "exon" features, and if it is a protein coding gene, a set of "CDS" (coding sequence) features. The CDS features cover a subset of the bases covered by the exon features. The "gene_id" and "transcript_id" columns let us know which transcript and gene each exon belongs to.

```
# --------------------------------------------------> gene
# 
# -------------------------------------------->       transcript
# ---------->         --->    ---------------->       exon
#       ---->         --->    ---------->             CDS
# 
# 
#                -----------------------------------> transcript
#                -------->       ---->    ----------> exon
#                     --->       ---->    -->         CDS
```

Let's look at this in the Integrative Genome Browser (IGV):

* Open IGV.
* Select the genome "C. elegans (ce11)" (top left drop-down box).
* Load the GTF file ("File" menu, "Load from file...").
* Search for "trx-1" in the location box (next to the "Go" button).

IGV turns all these features into a nice diagram of the gene's transcripts. The ENSEMBL gene annotation contains 5'UTR and 3'UTR regions, which are missing from the default gene annotations shown by IGV (probably from NCBI's RefSeq). `Gviz` or `ggbio` can be used to produce diagrams like this from within R (with a little work).



# Further operations on GRanges

💡 Using accessor methods like "start()" and "strand()" frees the authors of GRanges to store data internally using whatever efficiency tricks they want, and even to change this in a new version of Bioconductor. This is good, but it can be better:

💡 Double-stranded DNA has a rotational symmetry. None of the physics of DNA is changed if we look at it rotated end-to-end 180 degrees. Labelling of the strands as "+" and "-" is arbitrary. Also, it is usually not the exact position of features in a genome that is important so much as their relative position. If we restrict outselves to using functions that work just as well with either labelling of strands, and that are not affected by absolute position, our code will more directly express our intention and be less prone to bugs.

## Intra-range

Various useful manipulations of individual ranges are defined. 

```{r eval=FALSE}
?"intra-range-methods"
```

Note: How these make use of the strand is a little haphazard. For example `flank()`, `resize()`, and `promoters()` respect strand but `shift()` does not.

For example, suppose we are interested in the first 100 bases of a transcript:

```{r fig.height=1}
feat <- features[2]
feat_initial <- resize(feat, 100, fix="start")

look( c(feat, feat_initial) )
getSeq(seqs, feat_initial)
```

`resize` can fix either the `fix="start"` or `fix="end"` of the sequence.

`flank` can be either flank the start (`start=TRUE`) or end (`start=FALSE`).

```
#                                  5'     3'
# input                            ------->
#                                  .      .
# resize(input, n, fix="start")    -->    .
# resize(input, n, fix="end")      .    -->
#                                  .      .
# flank(input, n, start=TRUE)   -->.      .
# flank(input, n, start=FALSE)     .      .-->
#                                  .      .
# promoters(input, n1,n2)     -------->   .
#
```

## Inter-range

```{r eval=FALSE}
?"inter-range-methods"
?"nearest-methods"
?"setops-methods"
```

One compelling feature of GenomicRanges is that it is able to find overlapping ranges very quickly.

```{r}
query <- as("II:7752600-7754000:-", "GRanges")
hits <- findOverlaps(query, features, ignore.strand=TRUE)
hits
subjectHits(hits)
features[subjectHits(hits)]

findOverlaps(query, features, ignore.strand=FALSE)
```

With `findOverlaps`, we can use genomic location as the key when *joining* disparate types of data, so this is an important tool for *integrative* analysis. See also the related functions `nearest`, `precede`, `follow`, and `distance`.

`GenomicRanges` also provides:

* `range` - get a feature that spans from the start to the end of all features in a `GRanges`.
* `reduce` - merge overlapping features, so that the same bases are covered by a reduced collection of features.
* `disjoin` - as with reduce, but broken at each start and end of the input features.
* `setdiff` - subtracts one set of features from another, could be used with range on a set of exons to get introns. Might need to use `GenomicRanges::setdiff` if also using `dplyr`.

```
# input
#         --------->
#              ----------------->
#                      ------>
#                                     ---->
#                                     
# range   -------------------------------->
# 
# reduce  ----------------------->    ---->
# 
# disjoin ---->---->-->------>--->    ---->
# 
# setdiff(range(input),input)     --->
```

```{r fig.height=1}
trx1_exons <- subset(trx1_features, type == "exon")
look( trx1_exons )
look( range(trx1_exons) )
look( reduce(trx1_exons) )
look( disjoin(trx1_exons) )
```


## Challenge: transcript initiation {.challenge}

1. Make a GRanges of all transcripts with a biotype of "protein_coding".

2. Make a GRanges of the two bases immediately preceding each of these transcripts.

3. Get the DNA sequences of the ranges from part 2. Use `table()` to see if any sequences are particularly common.

Advanced: Get the four bases spanning the start of the transcript (two upstrand of the start and two downstrand of the start).

Note: In C. elegans transcript initiation is complicated by trans-splicing. I wonder what exactly the 5' ends of these transcript annotations represent?


# Example: Examining the PolyAdenylation Signal (PAS)

We'll now look at an example where the location of the thing we're looking for is more uncertain.

We suspect there is some motif that causes cleavage and polyadenylation of RNA transcripts, close to the 3' end of each transcript. Bioconductor will let us explore sequence around the ends of transcripts. It can also serve as to prepare data for command-line software such as the MEME Suite.

We'll limit our attention to protein coding transcripts.

```{r}
transcripts <- subset(features, 
    type == "transcript" & transcript_biotype == "protein_coding")

ends <- resize(transcripts, width=30, fix="end")
end_seqs <- getSeq(seqs, ends)
names(end_seqs) <- ends$transcript_id
```

We can check for general biasses in base composition:

```{r}
library(seqLogo)  

letter_counts <- consensusMatrix(end_seqs)

props <- prop.table(letter_counts[1:4,], 2)
seqLogo(props, ic.scale=FALSE)
seqLogo(props)
```

Saving the sequences to a file lets us apply command-line tools:

```{r}
writeXStringSet(end_seqs, "end_seqs.fasta")
```

This is suitable for use with, for example, the `streme` tool from MEME.

```
# streme --rna --minw 4 --maxw 15 --p end_seqs.fasta
```

Ok ok, so actually there is known PolyAdenylation Signal (PAS) motif, canonically "AAUAAA". Let's now look for this known motif.

```{r}
counts  <- vcountPattern("AATAAA", end_seqs)
table(counts)

matches <- vmatchPattern("AATAAA", end_seqs)
plot( tabulate(start(unlist(matches)),30), xlab="start position",ylab="matches")
```

We could also look for this pattern in the genome in general. Here is some code to find matches on either strand, using `vmatchPattern()`.

```{r}
query <- DNAString("AATAAA")
max.mismatch <- 0

fwd <- vmatchPattern(query, seqs, max.mismatch=max.mismatch)
fwd <- as(fwd, "GRanges")
strand(fwd) <- "+"

rev <- vmatchPattern(reverseComplement(query), seqs, max.mismatch=max.mismatch)
rev <- as(rev, "GRanges")
strand(rev) <- "-"

matches <- c(fwd, rev)
matches
```

Once we have a set of ranges, they can be related to other sets of ranges. For example, to reproduce the earlier result:

```{r}
overlaps <- findOverlaps(matches, ends, type="within")
table(table( subjectHits(overlaps) ))
```

The matches can be examined in IGV if saved in a GFF file. With a "tabix" index, IGV avoids the need to load the whole file.

```{r}
export(matches, "motif-matches.gff", index=TRUE)
```

The `pairwiseAlignment()` function provides more flexibility than `vmatchPattern()`, if needed. Command-line tools such as BLAST will be faster for longer query sequences.

Similar exploration can be performed around other regions of interest, such as peaks identified from ChIP. R may either provide a complete solution or serve as the glue plugging command-line tools together.



# BAMs and bigWigs

`import()` can read various other file types. This section will demonstrate reading read alignments from a BAM file and producing a bigWig depth of coverage file. The reads in the BAM file used here are a small sample from GSE57993 sample "N2 Rep1". https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE57993

Both BAM and bigWig files are designed so that specific regions can be accessed efficiently. Genome browsers can use this to only load data as needed. Bioconductor packages will also usually offer some way to load specific regions. This is useful because these files can be very large.

```{r}
alignments <- import("example.bam")
alignments

depth <- coverage(alignments)
depth

# Depth of coverage for gene rps-12
plot(depth$III[5679200:5680200], type="l")

export(depth, "depth.bw")
```

Explore the documentation for "GenomicAlignments", "Rle", and "RleList". "Rle" and "RleList" objects are memory-efficient vectors and lists of vectors. They support arithmetic operation such as scaling with `*` and adding with `+`. Rather neat! Sometimes a function won't support "Rle"s. When this happens, an Rle can be converted to a conventional vector with `as.numeric()`.

Filtered import of a BAM file can be performed with a suitable use of `ScanBamParam`, `import(bam, param=ScanBamParam( ... ))`:

* Specific strand.
* Specific region of reference genome.
* Specific set of single cells, based on CB attribute.

Try loading the "motif-matches.gff", "example.bam", and "depth.bw" files into IGV. Look for example at gene "rps-12".



# Transcript database objects

We've been using our genomic features as one big unstructured GRanges. This is messy. TxDb objects offer a more structured representation of genes, transcripts, exons, and coding sequences.

```{r}
library(GenomicFeatures)

txdb <- makeTxDbFromGRanges(features)
txdb
```

`txdb` is a "TxDb" object. Some ways to extract data from a TxDb are:

```{r results="hide"}
genes(txdb)
transcriptsBy(txdb, by="gene")
exonsBy(txdb, use.names=TRUE)
cdsBy(txdb, use.names=TRUE)

cds_ranges <- cdsBy(txdb, use.names=TRUE)
cds_ranges$B0228.5a.1
cds_ranges[["B0228.5a.1"]]
unlist(cds_ranges)
```

`cds_ranges` here is a "GRangesList". That is, a list containing GRanges objects. To get the transcript sequence or the coding sequence, these each need to be retrieved and then concatenated together. `extractTranscriptSeqs()` can do this for us.

```{r}
extractTranscriptSeqs(seqs, cds_ranges)
```

There's much more to explore here. Have a look at the documentation for the `GenomicFeatures` and `ensembldb` packages. To annotate a set of genomic features such as peaks from a ChIP-seq experiment, see for example the `ChIPseeker` package.


# Genome and annotation resources

(return to slideshow)

Besides software, Bioconductor includes various types of reference data.

## Example packages

* *BSgenome.Hsapiens.UCSC.hg38* Biostrings genome, Homo sapiens, from the UCSC browser, version hg38. DNA for chromosomes, usable in the same way as the DNAStringSet used above.

* *TxDb.Hsapiens.UCSC.hg38.knownGene* Transcript database, Homo sapiens, from UCSC browser, genome verison hg38, "knownGene" gene annotations. GRanges information for genes and transcripts, much as we loaded from a GTF file above.

* *org.Hs.eg.db* Organism Homo sapiens, primary key is Entrez Gene, database. Translation of gene ids from various databases, assignment to GO terms, KEGG pathways, etc. Entrez Gene ids are used as the primary key.

* *GO.db* GO Gene Ontology term descriptions and relationships. However to find which terms apply to specific genes, use the GOALL column in the relevant species' organism database.

Loading any of these packages gives you an `AnnotationDb` object (defined in the package `AnnotationDbi`). The main way to query these objects is with the `select()` function. `mapIds()` can be used if you need to enforce a one-to-one mapping.

```{r}
library(org.Ce.eg.db)

keytypes(org.Ce.eg.db)
columns(org.Ce.eg.db)

AnnotationDbi::select(org.Ce.eg.db,
    keys="trx-1", keytype="SYMBOL", columns=c("ENSEMBL","ENTREZID"))

head( AnnotationDbi::select(org.Ce.eg.db, 
    keys="WBGene00015062", keytype="ENSEMBL", columns="GOALL") )


library(GO.db)

AnnotationDbi::select(GO.db, 
    keys="GO:0005622", keytype="GOID", columns="TERM")
```

The TxDb we made earlier can also be used in this way.

```{r}
keytypes(txdb) 
columns(txdb)
```

Under the hood these are all SQLite databases. `select()` joins tables together as needed to answer a query.

```{r}
DBI::dbListTables( dbconn(txdb) )
DBI::dbListFields( dbconn(txdb), "transcript" )
```


## biomaRt 

[BioMart](http://www.biomart.org/) servers accessed using the `biomaRt` package are another way to get information such as translation of gene ids, gene sets, and gene information. For reproducibility, if using the ENSEMBL servers, make sure to specify a specific version of Ensembl to use. Your scripts may fail in future if the server you are using goes down or changes how it is accessed.

```{r eval=FALSE}
library(biomaRt)

mart <- useEnsembl(biomart = "genes", dataset = "celegans_gene_ensembl", version=104)

getBM(
    attributes=c("external_gene_name", "ensembl_gene_id", "entrezgene_id"), 
    filters="external_gene_name", values="trx-1", mart=mart)
```


## AnnotationHub

AnnotationHub is a way to retrieve data from a more comprehensive set of organisms and data providers than are provided as individual Bioconductor packages. The retrieved data is returned in an appropriate Bioconductor type. If data is being updated over time (eg improved annotation of a genome), each version receives a unique ID in AnnotationHub, making it possible to write reproducible analyses.

Files are cached, so they will only be downloaded once.

In the example below, the yeast genome and Ensembl annotations are retrieved:

```{r eval=FALSE}
library(AnnotationHub)
ah <- AnnotationHub()

# ah contains a large collection of records that can be retrieved
ah
length(ah)
colnames( mcols(ah) )
table( ah$rdataclass )

# There is an interactive Shiny search interface
display(ah)

# query() searches for terms in an unstructured way
query(ah, c("Ensembl", "96", "Saccharomyces cerevisiae"))

# Having located records of interest,
# your R script can refer to the specific AH... record,
# so it always uses the same version of the data.
sc_genome  <- ah[["AH70449"]]
sc_granges <- ah[["AH69700"]]
sc_txdb    <- ah[["AH69265"]]

# sc_genome is a TwoBitFile
# Can use getSeq on it without loading everything into memory
seqinfo(sc_genome)
getSeq(sc_genome, as("IV:1-10","GRanges"))
import(sc_genome)
```



# Package versions used

Use `sessionInfo()` to check the versions of packages used.

```{r}
sessionInfo()
```

[Home](index.html)
